---
title: "Affordability of Fresh Fruits in the U.S. in 2022: A Cup-Equivalent Price Analysis"
author: "Sarah Yao, Isabel Lange"
date: "`r Sys.Date()`"
output: github_document
---

## Introduction

The goal of this project is to analyze the affordability of fruits in the United States in 2022, focusing on their price per edible cup equivalent. Access to nutritious fruits is a key component of a healthy diet, yet prices vary widely depending on fruit type and form (fresh, frozen, canned, juice). Understanding which fruits are most and least affordable can help consumers make informed dietary choices and highlight potential areas for improving access to nutritious foods.

In pursuit of the stated goal, we will explore the following questions:

**1. Which fruits are most and least affordable per edible cup equivalent?**

We will sort all fruits by cup-equivalent price and identify the top 5 cheapest and most expensive fruits. A bar chart will visualize all fruits ranked by price, colored by fruit form, to contextualize affordability in terms of nutrition.

**2. How does fruit form (fresh vs. processed) affect price per cup?**

By grouping the data by form (Fresh, Dried, Canned, Frozen, Juice), we will compute mean, median, and standard deviation of cup-equivalent prices. A boxplot will compare price distributions across forms. This analysis will reveal whether fresh fruits tend to be cheaper or more expensive than processed forms.

**3. What is the relationship between fruit yield and cup-equivalent price?**

We will calculate the correlation between yield and cup-equivalent price, fit a linear regression model, and plot a scatterplot with a regression line. This will allow us to determine whether higher-yield fruits tend to be more affordable.

**4. What is the variability across fruit types in price per cup?**

Summary statistics including variance, standard deviation, range, and interquartile range will quantify price variability. A histogram  will visualize the spread, highlighting where most fruit prices cluster. This analysis will provide insights into consumer access and price stability across fruit categories.

Through the completion of this project, we aim to identify patterns in fruit affordability across the U.S. in 2022 and provide meaningful conclusions regarding the accessibility of nutritious foods for consumers.

## Data

### Structure

The link to the dataset is https://www.ers.usda.gov/data-products/fruit-and-vegetable-prices
. The USDA Economic Research Service (ERS) website provides datasets containing average retail prices of fruits and vegetables in the U.S. for selected years (2013, 2016, 2020, 2022). Each dataset contains multiple tables with price data for fresh, dried, frozen, canned, and juice forms of fruits and vegetables. For each fruit item, information includes retail price, yield, and cost per cup-equivalent. The datasets are compiled from retail scanner data provided by Circana/IRI across a variety of store types.

For this project, we are analyzing the dataset titled “ALL FRUITS – Average Prices” for the year 2022. This table contains 62 fruit items with data on retail price, yield, and cup-equivalent cost. Focusing on 2022 ensures that the data is recent, consistent in reporting methods, and allows us to study the current affordability of fruits in terms of nutrition-relevant units.

The dataset includes multiple forms of each fruit—fresh, canned, frozen, and juice. To manage the dataset and focus on meaningful analysis, we will include all 62 fruits but retain key variables: fruit name, form, retail price, yield, and cup-equivalent price. These variables provide enough information to analyze affordability, variability, and the relationship between fruit form and price. The dataset structure is as follows: each row corresponds to a unique fruit item in a specific form. The retail price column contains the average price in dollars, the yield column gives the fraction edible, and the cup-equivalent price column represents the cost per nutrition-relevant serving.

This structure allows us to perform analyses such as ranking fruits by cup-equivalent price, comparing prices across fruit forms, exploring correlations with yield, and quantifying variability across fruit types, while keeping the dataset manageable and focused on the core research questions.


### Cleaning 

First, we read in the CSV file for 2022.

```{r}
library(tidyverse)

fruit_2022 <- read_csv("Fruit-Prices-2022.csv")

```

Inspect the first few rows and column names

```{r}

head(fruit_2022)
colnames(fruit_2022)

```

Select only the relevant variables for analysis
Variables: Fruit name, Form, Retail Price, Yield, Cup-Equivalent Price

```{r}

fruit_2022 <- fruit_2022 %>% 
  select(Fruit, Form, RetailPrice, Yield, CupEquivalentPrice)


```

Check for missing values

```{r}

colSums(is.na(fruit_2022))

```

Ensure numeric columns are correctly formatted

```{r}

fruit_2022$RetailPrice <- as.numeric(fruit_2022$RetailPrice)
fruit_2022$Yield <- as.numeric(fruit_2022$Yield)
fruit_2022$CupEquivalentPrice <- as.numeric(fruit_2022$CupEquivalentPrice)

```

Check for duplicates

```{r}

sum(duplicated(fruit_2022))

```

Looks good! This dataset was pretty clean. We will save the cleaned dataset containing only the variables needed for this project to improve readability and make it easier to use in analysis.
``` {r}

write.csv(fruit_2022, "fruit_2022_clean.csv", row.names = FALSE)

```

### Variables

Fruit: The name of the fruit item (e.g., Apples, Strawberries, Oranges).

Form: The form in which the fruit is sold (Fresh, Frozen, Canned, Dried, or Juice).

RetailPrice: The average retail price of the fruit per standard unit in U.S. dollars.

Yield: The fraction of the fruit that is edible (e.g., 0.6 for 60% edible).

CupEquivalentPrice: The cost of one edible cup-equivalent of the fruit in U.S. dollars, calculated using retail price and yield.


## Results

```{r}

library(readr)
library(tidyverse)
clean <- read_csv("fruit_2022_clean.csv")

```
### Question 1: Which fruits are most and least affordable per edible cup equivalent?

To identify the most expensive fruits per edible cup equivalent, we will group fruits by their base names (for example, "Cherries" and "Cherries, packed in syrup or water" are grouped together) and calculate the average cup-equivalent price, ignoring the different forms for this analysis, as the goal is to identify the fruit itself.

```{r}
library(dplyr)
library(stringr)
library(purrr)

# Define your base fruits
base_fruits <- c("Apples", "Apricots", "Bananas", "Berries", "Blackberries", 
                 "Blueberries", "Cantaloupe", "Cherries", "Clementines", 
                 "Cranberries", "Dates", "Figs", "Fruit cocktail", "Grapefruit", 
                 "Grapes", "Honeydew", "Kiwi", "Mangoes", "Nectarines", "Oranges", 
                 "Papaya", "Peaches", "Pears", "Pineapple", "Plum", "Pomegranate", 
                 "Raspberries", "Strawberries")

# Create a FacetGroup column based on matching base fruits
clean <- clean %>%
  mutate(FacetGroup = map_chr(Fruit, ~ {
    matched <- base_fruits[str_detect(.x, base_fruits)]
    if(length(matched) > 0) matched[1] else .x
  }))

# Compute average cup-equivalent price per base fruit
clean_base <- clean %>%
  group_by(FacetGroup) %>%
  summarise(AvgPrice = mean(CupEquivalentPrice, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(AvgPrice)

# Top 5 cheapest base fruits
cheapest_fruits <- head(clean_base, 5)
cheapest_fruits

# Top 5 most expensive base fruits
most_expensive_fruits <- tail(clean_base, 5) %>%
  arrange(desc(AvgPrice))
most_expensive_fruits
```

Let's use a stacked bar graph to visualize the prices of each fruit across all of its forms, with each form distinguished by color.

```{r, fig.width=12, fig.height=20}

library(ggplot2)

# Summarize: total price per base fruit × form
stack_data <- clean %>%
  group_by(FacetGroup, Form) %>%
  summarise(TotalPrice = sum(CupEquivalentPrice, na.rm = TRUE), .groups = "drop")

# Reorder FacetGroup by total price across all forms
stack_data <- stack_data %>%
  group_by(FacetGroup) %>%
  mutate(GroupTotal = sum(TotalPrice)) %>%
  ungroup()

ggplot(stack_data,
       aes(x = reorder(FacetGroup, GroupTotal),
           y = TotalPrice,
           fill = Form)) +
  geom_col() +
  labs(
    title = "Stacked Bar Chart of Fruit Forms by Base Fruit",
    x = "Base Fruit",
    y = "Total Cup-Equivalent Price",
    fill = "Fruit Form"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold")
  )

```

Our graph confirms that, based on the average price per cup equivalent, watermelon is the cheapest fruit, while cherries are the most expensive.

Watermelon, along with common fruits such as apples and bananas, falls toward the lower end of the price range. In contrast, fruits like cherries, apricots, and raspberries are more expensive because they are labor-intensive to harvest, have short growing seasons, and spoil quickly, all of which increase production and handling costs. Their sensitivity to weather and high risk of crop loss further raise costs. These differences in production scale, labor demands, and seasonal constraints explain why watermelon remains the most affordable fruit per cup equivalent, while cherries are the most expensive.

Among the more expensive fruits, the canned form generally accounts for the majority of their cost, due to processing, packaging, and added preservatives. Raspberries are an exception, with their fresh form representing most of their price, likely because of their delicate nature, short shelf life, and labor-intensive harvesting.


### Question 2: How does fruit form (fresh vs. processed) affect price per cup?

To examine the price distribution of fruits across their different forms, we group the data by fruit form and compute the mean, median, and standard deviation of cup-equivalent prices. The boxplot below visualizes how fresh fruits compare with processed forms such as canned, frozen, and juice.

```{r}

# Summary statistics by fruit form
form_summary <- clean %>%
  group_by(Form) %>%
  summarise(
    MeanPrice   = mean(CupEquivalentPrice, na.rm = TRUE),
    MedianPrice = median(CupEquivalentPrice, na.rm = TRUE),
    SDPrice     = sd(CupEquivalentPrice, na.rm = TRUE),
    Count       = n()
  ) %>%
  arrange(MeanPrice)

form_summary

```

From this table, we see that when comparing mean and median cup-equivalent prices, juice is the cheapest fruit form, while frozen fruit is the most expensive. A chart below will help further illustrate this pattern for analysis.

```{r}

# Boxplot comparing price distributions across forms
ggplot(clean, aes(
  x = reorder(Form, CupEquivalentPrice, FUN = median),
  y = CupEquivalentPrice,
  fill = reorder(Form, CupEquivalentPrice, FUN = median)
)) +
  geom_boxplot(outlier.alpha = 0.6) +
  labs(
    title = "Price per Cup Equivalent by Fruit Form",
    x = "Fruit Form",
    y = "Cup-Equivalent Price ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
    plot.title = element_text(size = 14, face = "bold")
  ) +
  guides(fill = "none")

```

The boxplot reveals clear differences in price per cup equivalent across fruit forms. 

Juice is the least expensive overall, with a tight distribution and no extreme outliers, likely because juice production benefits from large-scale manufacturing, blending, and dilution, which lower the cost per edible cup. In contrast, frozen fruit shows the highest median price and a wider spread, indicating greater variability in production and storage costs. This is interesting because we originally predicted that fresh fruit would be the most expensive. However, further research revealed that frozen fruits require additional processing steps such as washing, cutting, and flash-freezing, as well as energy-intensive cold storage and transportation, all of which increase cost. Fresh fruits remain relatively moderate in price, and the outlier in this category likely reflects specialty produce—fruits that are highly perishable, seasonal, or labor-intensive. Canned fruits sit in the middle, but the presence of a high outlier suggests that certain canned varieties can be significantly more expensive due to added sugars, packaging differences, or supply limitations. Overall, the variability and outliers across forms highlight how processing intensity, storage methods, and supply-chain complexity shape the cost of a single cup-equivalent of fruit.

### Question 3: What is the relationship between fruit yield and cup-equivalent price?

```{r}
# Correlation between yield and cup-equivilance price
correlation <- cor(clean$Yield, clean$CupEquivalentPrice, use = "complete.obs")
correlation
```

The correlation between yield and cup-equivalent price is −0.0469, indicating a very weak negative relationship. While the negative sign suggests that fruits with higher yields tend to have slightly lower cup-equivalent prices, the magnitude is very close to zero, meaning that yield explains almost none of the variation in price.

```{r}
# Linear regression model
model <- lm(CupEquivalentPrice ~ Yield, data = clean)
summary(model)
```

The slope of the regression model comes from the row labeled “Yield” in the coefficients table. In a linear regression output, each predictor has its own row, and the Estimate column provides the slope, which represents the predicted change in the response variable for a one-unit increase in the predictor. Since Yield is the predictor in this model, its estimate of –0.1551 is the slope. This means the model predicts that cup-equivalent price decreases by about 0.16 dollars for each one-unit increase in yield.

However, the negative slope is not statistically significant. A high p-value of 0.7172 indicates that the observed slope could easily be due to random variation rather than a true effect. In addition, the model’s R squared value is extremely low (0.22 percent), showing that yield explains almost none of the variation in cup-equivalent prices. Although the slope suggests a slight downward trend, the effect is weak, unreliable, and not practically meaningful. Overall, the regression analysis demonstrates that yield is not a useful predictor of cup-equivalent price.

```{r}
library(ggplot2)

# Scatterplot of the two variables, containing a regression line
<<<<<<< HEAD
ggplot(clean, aes(x = Yield, y = CupEquivalentPrice, col=Fruit)) +
=======
ggplot(clean, aes(x = Yield, y = CupEquivalentPrice, color = FacetGroup)) +
>>>>>>> f5261648a7b66a38442b491f95cb230c08d6739a
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  labs(
    title = "Relationship Between Fruit Yield and Cup-Equivalent Price",
    x = "Yield (fraction edible)",
    y = "Cup-Equivalent Price ($)",
    color = "Fruit"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold")
  )

```

The scatterplot shows a large amount of variability and no clear trend. The regression line is nearly flat, which visually supports the statistical results: there is no strong relationship between yield and price.


### Question 4: What is the variability across fruit types in price per cup?

```{r}
# Summary Statistics
variance <- var(clean$CupEquivalentPrice, na.rm = TRUE)
std_dev  <- sd(clean$CupEquivalentPrice, na.rm = TRUE)
range_vals <- range(clean$CupEquivalentPrice, na.rm = TRUE)
iqr_val <- IQR(clean$CupEquivalentPrice, na.rm = TRUE)

list(
  variance = variance,
  standard_deviation = std_dev,
  range = range_vals,
  IQR = iqr_val
)
```

Summary statistics and visualizations depict the fruit prices per cup variability. The variance (0.334) and standard deviation (0.578) indicate some spread around the mean price. Prices range from $0.24 to $3.56 per cup.

```{r}
library(ggplot2)

# Histogram
ggplot(clean, aes(x = CupEquivalentPrice)) +
  geom_histogram(bins = 15, fill = "blue", color = "black") +
  labs(
    title = "Distribution of Fruit Prices per Cup",
    x = "Cup-Equivalent Price ($)",
    y = "Count"
  )

# Density Plot
ggplot(clean, aes(x = CupEquivalentPrice)) +
  geom_density(fill = "green", alpha = 0.5) +
  labs(
    title = "Density of Fruit Cup-Equivalent Prices",
    x = "Cup-Equivalent Price ($)",
    y = "Density"
  )
```

The histogram and density plot show that fruit prices are right-skewed, with most fruits clustered at relatively low prices between about $0.20 and $1.75 per cup equivalent. The pronounced peak around $1.00 suggests this is a common price point for many widely available fruits, while the density gradually tapers off toward higher prices. Only a small number of fruits fall above $2.00, forming a long upper tail that likely reflects specialty, imported, or out-of-season options. This distribution indicates that while most fruits are generally affordable, access to higher-variety or premium fruits comes at a steeper cost, creating meaningful differences in affordability across types.

Together, these statistics and visualizations indicate that price per cup varies considerably across fruit types, which could affect consumer access depending on the fruit.

## Conclusion

This analysis of fruit affordability in the United States in 2022 reveals several important insights that challenge common assumptions about nutritious eating on a budget. First, affordability varies widely across fruit types, with most fruits clustering around 1.00 dollars per cup equivalent but a long tail of higher priced fruits, particularly berries and cherries, driving substantial variability. Watermelon stands out as the most affordable option, while cherries are the most expensive, reflecting differences in seasonality, labor intensity, and handling requirements.

Our investigation into fruit forms shows that processing plays a significant role in pricing. Contrary to the expectation that fresh fruits would be the most expensive, frozen fruits had the highest median price and the widest variability, likely due to their additional processing and storage demands. Juice was the least expensive and most consistent form, suggesting that large scale processing and blending help reduce costs. These findings highlight the importance of considering not only what fruit to buy but also the form in which it is purchased.

We also examined whether yield, the fraction of a fruit that is edible, influences price. Both the correlation and regression analyses indicate that yield has almost no relationship with cup equivalent price. This demonstrates that other factors, such as supply chain complexity, perishability, and processing, are far more influential in determining affordability.

Finally, the substantial variability across fruit types suggests that while many fruits are accessible at low cost, others remain significantly more expensive, which may limit consumer access to certain varieties. Future research incorporating different years, allowing for more comparison, could help identify consistent patterns in fruit pricing over time. Examining seasonal trends could provide further insight into how consumers can maximize affordability while maintaining a nutritious diet.
